\sysfull

### 2018-08-04 paper/asplos19/paging.tex
restructure the paper directory
Figure 6 updated and explained through section 5

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figures/graffle/paging.pdf}
	\caption{Illustration of task-virtual memory interaction in \sys. Tasks only interact with the working buffer in SRAM (indicated as page buffer). The shadow buffer is used to hold the dirty bages. The indirection table is used to speed-up two-phase commit operation: it eliminates the copy operation in the second phase.}
	\label{figure:coala_page_tags}
\end{figure}

\textbf{System Overview.} Figure~\ref{figure:coala_page_tags} provides an
overall view of \sys's virtual memory management system.  A page has a home
location in non-volatile memory which we refer to as the page's
\texttt{\underline{store}} (underlined locations are non-volatile).  When a
task accesses a page, the memory manager swaps the page from its home location
into the fully\hyp{}associative \texttt{working} buffer that is stored in
volatile memory.  The \texttt{working} buffer is allocated to take up the
volatile memory that remains after space for stack is allocated.
%
\sys uses a two-phase commit to write dirty pages from the \texttt{working}
buffer back to their home location in \texttt{\underline{store}} atomically,
ensuring data consistency even in the presence of a power failure during commit.
%
The memory manager is forced to swap out a dirty page from the \texttt{working} buffer 
when a task accesses a page that is not in the working buffer, and there is no
free page in the working buffer (i.e., a page fault). On a page fault, \sys
moves the page to a direct\hyp{}mapped non-volatile \texttt{\underline{shadow}}
page buffer. Later, to commit the page, \sys writes it from its
\texttt{\underline{shadow}} location back to its home location in
\texttt{\underline{store}}.
%
%BML: this point is true, but without caches, it is not a performance factor, so I removed it.
%Unlike prior work~\cite{chain,alpaca} that moves scattered variables one at a
%time between volatile and non-volatile memory, \sys's page-based design moves
%contiguous blocks of data.
%
%BML: this seems gratuitous because DMA is very common and we've already introduced it.
%\sys takes full advantage of hardware\hyp{}accelerated block copies using
%Direct Memory Access (DMA) hardware in common microcontrollers --- e.g.
%5$\times$ fewer cycles on MSP430FRxxxx.

@@

A \sys task must access protected non-volatile variables through \sys's
restricted memory interface, which Section~\ref{sec:coala_api} describes in
detail.  The interface includes \texttt{RP(var)}, to read the value of variable
{\tt var}, and \texttt{WP(var)} to assign a value to variable {\tt var}. The
implementation of {\tt RP} is shown in Algorithm~\ref{algo:rwar}, and {\tt
WP}'s implementation is similar except that {\tt WP} marks the accessed page as
dirty.
%
{\tt RP} and {\tt WP} operations translate a variable's physical address in
non-volatile memory into a \emph{virtual address} in the page buffer in
volatile memory. In \sys, a virtual address is composed of a \emph{page tag}
that identifies the page and a \emph{page offset} that identifies a byte.

@@

%\begin{algorithm}[t]
\begin{wrapfigure}{t}{0.4\textwidth}
\fbox{\parbox{\linewidth}{
	\captionof{algorithm}{\texttt{RP}(variable $v$)}
	\label{algo:rwar}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
		\State $t \leftarrow \Call{GetTag}{v}$ 
        \State $i \gets \{ j\ |\ \Call{GetTag}{\texttt{working}[j]} = t \}$ \Comment{Search page}
		\If { $i = \emptyset$ } \Comment{Var in a resident page?}
		\State	$i \gets \Call{PageFault}{t}$ \Comment{Swap page in}
		\EndIf
		\State $o \leftarrow \Call{GetOffset}{v}$ 		
		\State \Return \texttt{working}[$i$][$o$]  \Comment{Return from page}
%	\end{algorithmic}
%\end{algorithm}
	\end{algorithmic}}}
\end{wrapfigure}

After address translation, \sys attempts to access the protected variable's
location in the volatile paging buffer. \sys keeps track of the page tags for
the pages currently resident in the paging buffer. When a task accesses a
variable, it compares the variable's page tag to tags of the pages in {\tt
working} (Line 2).
%
%If the accessed variable's page is resident in the page buffer,
%then the tag of one of the pages in this buffer is equal to the accessed
%variable's page tag.
%
If the accessed variable's page tag is not found in the page buffer, the
operation incurs a {\em page fault} (Line 4). At a page fault, \sys must swap
out one of the pages that is resident in the page buffer (to the non-volatile
\texttt{\underline{shadow}}) if \texttt{working} is full, and swap in the
variable's page.
%
The byte is accessed in the page buffer at the index of the resident page and
the variable's page offset (Lines 5--6).

@@

If the \texttt{working} buffer is full, a page fault on a memory access
requires \sys to swap out one of the pages in the volatile buffer (a 
\emph{victim page}) preserving updates made to that page, and to copy the
accessed page of protected data into the working buffer.
Algorithm~\ref{algo:pagefault} shows how \sys handles a page fault. A victim
page in the \texttt{working} buffer is selected for eviction according to
a \emph{first-in-first-out} (FIFO) replacement policy (Lines 2 and 12). 
%
While we opted for the simplicity of FIFO, any other replacement policy, such
as Least Recently Used (LRU) could work in its place.

% \begin{algorithm}
% \begin{wrapfigure}{t}{0.5\textwidth}
% \fbox{\parbox{}{
% 	\captionof{algorithm}{\texttt{PageFault}(tag $t$)}
% 	\label{algo:pagefault}
% 	\scriptsize
\begin{algorithm}
	\caption{\texttt{PageFault}(tag $t$)}
	\label{algo:pagefault}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
        \State \LeftComment{$i_\text{victim}$: index to free slot or page to be evicted, initially 0}
        \State \LeftComment{\texttt{\underline{shadowCount}}: non-volatile, 0 on boot unless in phase 2 commit}
        \State $t_\text{victim} \gets \Call{Tag}{\texttt{working}[i_\text{victim}]}$
		\If {\Call{isDirty}{working[$i_\text{victim}$]} }	\Comment{Evicting a dirty page?}
		\State $\texttt{\underline{shadow}}[t_\text{victim}] \xleftarrow{\text{DMA}} \texttt{working}[i_\text{victim}]$
            \Comment{Swap out the page}
        \If {$t_\text{victim} \not\in \texttt{\underline{shadowList}}$} \Comment{Keep track for quick traversal}
            \State $\texttt{\underline{shadowList}}[\texttt{shadowCount++}] \gets t_\text{victim}$
        \EndIf
		\EndIf
		\If {\Call{exists}{\underline{shadow}[$t$]}} \Comment{Page swapped out earlier?}
		\State $\texttt{working}[i_\text{victim}] \xleftarrow{\text{DMA}} \texttt{\underline{shadow}}[t]$
            \Comment{Get from commit buffer}
		\Else
		\State $\texttt{working}[i_\text{victim}] \xleftarrow{\text{DMA}} \texttt{\underline{store}}[t]$
            \Comment{Get from backing store}
		\EndIf 
        \State $i_\text{victim} \gets (i_\text{victim} + 1) \mod |\texttt{pageBufLength}|$ \Comment{FIFO evict}
	\end{algorithmic}
\end{algorithm}

@@

If a task modified any byte in the victim page (i.e., using \texttt{WP}), then
the page is dirty.  If the page being swapped out is not dirty and the page
being accessed had not been dirtied and swapped out to the \texttt{shadow}
buffer since the last reboot, then handling a page fault requires copying the
accessed page from its home in the  \texttt{\underline{store}} into the
\texttt{working} buffer (Line 11).  Otherwise, the page fault handler must take
the following extra steps.
%
Before swapping in the accessed page, \sys must first preserve the updated
values in the dirty page so that they can be committed back to their locations
in memory when the task commits (but not earlier).  To preserve the dirty
values, \sys copies the victim page to the non-volatile
\texttt{\underline{shadow}} buffer (Line 5) and records the page's tag in a
non-volatile commit list (Lines 6-7). \sys uses the commit list to iterate over shadow pages 
at commit time, to copy the contents of shadow pages back to the main memory {\tt store}. 
%
%Dirty pages remain in {\tt \underline{shadow}} buffer until the
%dynamic task completes, at which point \sys commits the modified
%pages from {\tt \underline{shadow}} back to \texttt{\underline{store}}.
%
If the accessed page was previously modified and swapped out since the last
power failure (Line 8), the most recent version of the page is in {\tt
\underline{shadow}}, not \texttt{\underline{store}} (Line 9).

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/mem-man.pdf}
    \caption{Interaction between task, memory manager and memory buffers upon \texttt{RP} and \texttt{WP},
    and order of occurrence. Steps 2, 3, 4 and 5 are conditional. Tasks only have direct access to the
    \texttt{working} buffer.}
    \label{figure:mem-man}
\end{figure}

% \begin{figure}
	% \centering
	% \includegraphics[width=\columnwidth]{figures/paging.pdf}
	% \caption{Illustration of task-virtual memory interaction in \sys. Tasks only interact with the working buffer in SRAM (indicated as page buffer). The shadow buffer is used to hold the dirty bages. The indirection table is used to speed-up two-phase commit operation: it eliminates the copy operation in the second phase.}
	% \label{figure:coala_page_tags}
% \end{figure}

\textbf{System Overview.}
Figure~\ref{figure:mem-man} gives an overview of \sys's virtual memory management system,
whose details are provided throughout this section.
A page has a home location in non-volatile memory which we refer to as the page's
\texttt{\underline{store}} (underlined locations are non-volatile).  When a
task accesses a page, the memory manager swaps the page from its home location
into the fully\hyp{}associative \texttt{working} buffer that is stored in
volatile memory.  The \texttt{working} buffer is allocated to take up the
volatile memory that remains after space for stack is allocated.
%
\sys uses a two-phase commit to write dirty pages from the \texttt{working}
buffer back to their home location in \texttt{\underline{store}} atomically,
ensuring data consistency even in the presence of a power failure during commit.
%
The memory manager is forced to swap out a dirty page from the \texttt{working} buffer
when a task accesses a page that is not in the working buffer, and there is no
free page in the working buffer (i.e., a page fault). On a page fault, \sys
moves the page to a direct\hyp{}mapped non-volatile \texttt{\underline{shadow}}
page buffer. Later, to commit the page, \sys writes it from its
\texttt{\underline{shadow}} location back to its home location in
\texttt{\underline{store}}.
%
%BML: this point is true, but without caches, it is not a performance factor, so I removed it.
%Unlike prior work~\cite{chain,alpaca} that moves scattered variables one at a
%time between volatile and non-volatile memory, \sys's page-based design moves
%contiguous blocks of data.
%
%BML: this seems gratuitous because DMA is very common and we've already introduced it.
%\sys takes full advantage of hardware\hyp{}accelerated block copies using
%Direct Memory Access (DMA) hardware in common microcontrollers --- e.g.
%5$\times$ fewer cycles on MSP430FRxxxx.

@@

A task must access protected non-volatile variables through \sys's
restricted memory interface. The interface includes \texttt{RP}$(v)$, to read the value of variable
$v$, and \texttt{WP}$(v)$ to assign a value to variable $v$. The
implementation of {\tt RP} is shown in Algorithm~\ref{algo:rwar}, and {\tt
WP}'s implementation is similar except that {\tt WP} marks the accessed page as
dirty.
%
{\tt RP} and {\tt WP} operations translate a variable's physical address in
non-volatile memory into a \emph{virtual address} in the page buffer in
volatile memory. In \sys, a virtual address is composed of a \emph{page tag}
that identifies the page and a \emph{page offset} that identifies a byte.

@@

%\begin{algorithm}[t]
\begin{figure} %{t}{0.4\textwidth}
\fbox{\parbox{\linewidth}{
	\captionof{algorithm}{\texttt{RP}(variable $v$)}
	\label{algo:rwar}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
		\State $t \leftarrow \Call{GetTag}{v}$
        \State $i \gets \{ j\ |\ \Call{GetTag}{\texttt{working}[j]} = t \}$ \Comment{Search page}
		\If { $i = \emptyset$ } \Comment{Var in a resident page?}
		\State	$i \gets \Call{PageFault}{t}$ \Comment{Swap page in}
		\EndIf
		\State $o \leftarrow \Call{GetOffset}{v}$
		\State \Return \texttt{working}[$i$][$o$]  \Comment{Return from page}
%	\end{algorithmic}
%\end{algorithm}
	\end{algorithmic}}}
\end{figure}

After address translation (Figure~\ref{figure:mem-man}, Step 6), a task
accesses the protected variable's location in the volatile paging buffer
(Step 7). \sys keeps track of the page tags for
the pages currently resident in the paging buffer. When a task accesses a
variable, it compares the variable's page tag to tags of the pages in {\tt
working} (Algorithm~\ref{algo:rwar}, Line 2).
%
%If the accessed variable's page is resident in the page buffer,
%then the tag of one of the pages in this buffer is equal to the accessed
%variable's page tag.
%
If the accessed variable's page tag is not found in the page buffer, the
operation incurs a {\em page fault} (Line 4).
% At a page fault, \sys must swap
% out one of the pages that is resident in the page buffer (to the non-volatile
% \texttt{\underline{shadow}}) if \texttt{working} is full, and swap in the
% variable's page.
%
The byte is accessed in the page buffer at the index of the resident page and
the variable's page offset (Lines 5--6).

@@

When accessing a protected variable with \texttt{RP} or \texttt{WP}, the memory
manager first searched the variable's page in the \texttt{working} buffer
(Figure~\ref{figure:mem-man}, Step 1).
If the page is not found there, a page fault is incurred and a new page needs to
be swapped in.
If the \texttt{working} buffer is full, a page fault on a memory access
requires \sys to swap out one of the pages in the volatile buffer (a
\emph{victim page}) preserving updates made to that page.
% and to copy the accessed page of protected data into the working buffer.
% Algorithm~\ref{algo:pagefault} shows how \sys handles a page fault. A victim
% page in the \texttt{working} buffer is selected for eviction according to
% a \emph{first-in-first-out} (FIFO) replacement policy (Lines 2 and 12).
%
While we opted for the simplicity of FIFO, any other replacement policy, such
as Least Recently Used (LRU) could work in its place.

% \begin{algorithm}
% \begin{wrapfigure}{t}{0.5\textwidth}
% \fbox{\parbox{}{
% 	\captionof{algorithm}{\texttt{PageFault}(tag $t$)}
% 	\label{algo:pagefault}
% 	\scriptsize
% \begin{algorithm}
	% \caption{\texttt{PageFault}(tag $t$)}
	% \label{algo:pagefault}
	% \scriptsize
	% %\small
	% \begin{algorithmic}[1]
        % \State \LeftComment{$i_\text{victim}$: index to free slot or page to be evicted, initially 0}
        % \State \LeftComment{\texttt{\underline{shadowCount}}: non-volatile, 0 on boot unless in phase 2 commit}
        % \State $t_\text{victim} \gets \Call{Tag}{\texttt{working}[i_\text{victim}]}$
		% \If {\Call{isDirty}{working[$i_\text{victim}$]} }	\Comment{Evicting a dirty page?}
		% \State $\texttt{\underline{shadow}}[t_\text{victim}] \xleftarrow{\text{DMA}} \texttt{working}[i_\text{victim}]$
            % \Comment{Swap out the page}
        % \If {$t_\text{victim} \not\in \texttt{\underline{shadowList}}$} \Comment{Keep track for quick traversal}
            % \State $\texttt{\underline{shadowList}}[\texttt{shadowCount++}] \gets t_\text{victim}$
        % \EndIf
		% \EndIf
		% \If {\Call{exists}{\underline{shadow}[$t$]}} \Comment{Page swapped out earlier?}
		% \State $\texttt{working}[i_\text{victim}] \xleftarrow{\text{DMA}} \texttt{\underline{shadow}}[t]$
            % \Comment{Get from commit buffer}
		% \Else
		% \State $\texttt{working}[i_\text{victim}] \xleftarrow{\text{DMA}} \texttt{\underline{store}}[t]$
            % \Comment{Get from backing store}
		% \EndIf
        % \State $i_\text{victim} \gets (i_\text{victim} + 1) \mod |\texttt{pageBufLength}|$ \Comment{FIFO evict}
	% \end{algorithmic}
% \end{algorithm}

@@

If a task modified any byte in the victim page (i.e., using \texttt{WP}), then
the page is dirty and its changes has to be persisted to non-volatile memory
(Figure~\ref{figure:mem-man}, Step 2).
If the accessed page was previously modified and swapped out since the last
power failure, the most recent version of the page is in {\tt
\underline{shadow}} (Step 3), otherwise it has to be retrieved from
{\tt \underline{store}} (Step 4).
Finally, the page is copied to the volatile buffer (Step 5).
%
%Dirty pages remain in {\tt \underline{shadow}} buffer until the
%dynamic task completes, at which point \sys commits the modified
%pages from {\tt \underline{shadow}} back to \texttt{\underline{store}}.
%
